Epistemic Semantics in Guarded String Models

### 2020-09-25 tex/sdraft.tex
prose tightening in 4 - 6

\newcommand{\ehc}[1]{\textit{\textcolor{red}{[ehc]: #1}}}
\newcommand{\mr}[1]{\textit{\textcolor{cyan}{[mr]: #1}}}

@@

\begin{hangingpar}
{\tt{define Kst(R)
    [St\hspace{1pt}.x.\hspace{1pt}St]\hspace{1pt}|\hspace{1pt}Kpl(X);}}\\ The Fst
operation {\tt{.x.}} is Cartesian product.  {\tt{R.i}} is the inverse of relation
{\tt{R}}.
\end{hangingpar}
\end{tcolorbox}
\caption{Definition in Fst of the Kleene concatenation closure of
  a relation between guarded strings.}
\label{fstfigure2}
\end{figure}

@@

In the construction in Fst, the definition of world alternatives takes
a simple form. Where $R_a$ is the relation on decorated events for
agent $a$, the corresponding relation on worlds in is the Kleene
closure of $R_a$.  Where $R$ and $S$ are relations, the concatenation
product of $R$ and $S$ is the set of pairs of the form
$\pair{x_1x_2}{y_1y_2}$, where $\pair{x_1}{y_1}$ is in relation $R$,
and $\pair{x_2}{y_2}$ is in relation $S$.  The Kleene closure of
relation $R$ is \(\cup_{n \geq 0}R^n\), where $R^n$ is the $n$-times
concatenation product of $R$ with itself (the $0$-times concatenation
product is $\sem{ 1 }^{\varphi,\mathcal E}$). This is an operation in
the finite state calculus.  Figure \ref{relclosure} defines the
corresponding operation in the guarded string algebra.  The epistemic
alternative relation on worlds for an agent is then defined as the
concatenation closure of the event alternative relation for the agent.

@@

We also implement the semantics of Epik terms using lazy
lists~\cite{LazyLists} in Haskell, rather than the direct
interpretation as sets. Unfortunately, regular expressions, and hence
also Epik, can denote infinite sets of strings, for example the term
$\bullet^*$. Normally regular languages are represented using a finite
coalgebra~\cite{}. However the non-distributivity of $\diamond$
accross $;$ complicates this construction\footnote{Axiomatic and
  Coalgebraic models for Epik are open questions}. To sidestep this,
we parameterize the interpretation function on a positive integer $n$
and only produce guarded strings of length $n$ or less.

We translate the Epik terms into a Haskell algebraic datatype that
represents terms with the same signature as described in Section 2. We
then paramaterize the interpretation function on an integer $n$ that
describes the maximum length of a string we will produce.

The bounded interpretation into lists of strings is very similar to the
unbounded interpretation into sets of strings, except for the bounds
checking. The full details are shown in Figure~\ref{fig:list-interp}. First note
that when $n=0$, we simply return the empty list, denoted $[]$. Terms of the
form $0$, $1$, $e$, and $\psi$ have the same denotation as before, translated
into a list (for a set $X$, $\lfloor X \rfloor$ is a list with the same elements
as $X$ arbitrarily ordered). We compute these lists using BBDs, a standard
technique for concisely and cannoncially representing boolean functions~\cite{BBDS}.

We lift the remaining operators (except Kleene star) to their list
equivalents: union becomes list append (written $\texttt{+\!+}$);
concatenation becomes the fusion product (lifted to lists this time),
negation is implemented using list difference ($\setminus$), and the
modal operator lifts the alternative relation over lists of
strings\footnote{Figure~\ref{fig:list-interp} depicts this using the
  list comprehension notation, which is analogous to set builder
  notation, except that it is written using square brackets. Element
  order is evoked by the keyword \texttt{for}, rather than using the
  unordered $\forall$.}. The only caveat to these direct
interpretations is that we restrict the operators to have size $\leq
n$, denoted as $l|_n$ for a list of guarded strings $l$.

@@

The lists we use here are \emph{lazy} (as opposed to \emph{strict}), which
broadly means that computation is delayed until the value is needed. This allows
us to avoid computing large, unnecessary iterations.

@@

Sentences are parsed with a chart parser for categorial grammar.  The
semantics for complex phrases are obtained by application of semantic
translations, accompanied by beta reductions that eliminate all
lambdas in logical forms for clauses.  In consequence, the semantic
term translating a sentence is a term of the logical language.  Such a
term designates a set of possible words (guarded strings) in the
possible worlds model determined by an Epik specification.  By way of
example, \exi{three}{a} is an English sentence with conjunction and
several levels of clausal embedding.  Using the grammar and parser,
the sentence is mapped to the term in \exi{three}{b}. Then
\exi{three}{c} shows a simplified logical from constructed form
\exi{three}{b} using simply logical equivalences. The negations are
propositional. The double negation, while logically redundant, comes
syntactically from the grammar. This term is compiled in an
implementation of the finite state calculus to a finite state machine
with six states and eight arcs, which accepts a countably infinite set
of worlds.  In this way the methodology ``directly'' represents the
set of worlds denoted by \exi{three}{a}.

@@

\begin{examples} \label{three}
\itema It's tails and Amy knows that Bob knows that Amy knows whether it's heads.
% (_*;t) & ~ <amy> ~ ~ <bob> ~((~ (<amy> ~ (_*;h))) + ~ (<amy> (_*;h)) )
\itemb \( \displaystyle \bullet^*t \wedge \AKBKAKWH \)\\
\itemc \(\bullet^*t \wedge \K_{a}(\K_{b}(\K_{a} \bullet^* h + \K_{a} \neg \bullet ^* h))\)
\itemd Amy knows that it's tails.
\iteme \(
\bullet ^*t \wedge \neg \Diamond_{a} \neg \bullet^*t~~~~~~(\equiv \K_a \bullet^* t)
\)

@@

\section{Discussion}
The methodology presented here is designed for use in research in
linguistic semantics, and for education at the level of a second
graduate course in formal semantics, covering intensionality.  There
are straighforward extensions to additional semantic phenomena, such
as tense and perfective aspect as in \exi{extension}{a}, and the combination of
metaphysical modality and prospective aspect in \exi{extension}{b}.

>>>>

\newcommand{\ehc}[1]{\textit{\textcolor{red}{[ehc]: #1}}}
\newcommand{\mr}[1]{\textit{\textcolor{cyan}{[mr]: #1}}}
%% \newcommand{\ehc}[1]{}
%% \newcommand{\mr}[1]{}

@@

\begin{hangingpar}
{\tt{define RelKst(R)
    [St\hspace{1pt}.x.\hspace{1pt}St]\hspace{1pt}|\hspace{1pt}Kpl(X);}}\\ The Fst
operation {\tt{.x.}} is Cartesian product.  {\tt{R.i}} is the inverse of relation
{\tt{R}}.
\end{hangingpar}
\end{tcolorbox}
\caption{Definition in Fst of the Kleene concatenation closure of
  a relation between guarded strings.}
\label{fstfigure2}
\end{figure}

@@

In Fst, we use the Kleene closure of the concatenation product
operation on relations to lift a relation on events for an agent $a$,
$R_a$, to the corresponding relation on worlds.  First define the
concatenation product $R~S$ of two relations $R$ and $S$ to be the set
of pairs of the form $\pair{x_1x_2}{y_1y_2}$, where $x_1 \mathrel R
y_1$, and $x_2 \mathrel S y_2$. Then $R+$ is the closure of $R$ with
respect to this operation. Figure \ref{relclosure} defines the
corresponding operation in the guarded string algebra.  The epistemic
alternative relation on worlds for an agent is then defined as the
concatenation closure of the event alternative relation for the agent.

@@

We also implement the semantics of Epik terms using lazy lists in
Haskell, rather than the direct interpretation as sets. Using lists
sidesteps checking the set invariant (elements are unique) for large
sets, such as $\bullet^*$, and laziness allows us to delay computing
these large sets until they are actually needed. To sidestep the
infiteness of models, we parameterize the interpretation function on a
positive integer $n$ and only produce guarded strings of length $n$ or
less.

%% We translate the Epik terms into a Haskell algebraic datatype that
%% represents terms with the same signature as described in Section 2. We
%% then paramaterize the interpretation function on an integer $n$ that
%% describes the maximum length of a string we will produce.

The bounded interpretation into lists of strings is very similar to
the unboundeds interpretation into sets of strings, except for the
bounds checking. The full details are shown in
Figure~\ref{fig:list-interp}. First note that when $n=0$, the
denotation is empty, denoted $[]$. Terms of the form $0$, $1$, $e$,
and $\psi$ have the same denotation as before, translated into a list
(denoted $\lfloor S\rfloor$, for a set $S$). We compute atoms using
BBDs, which concisely represent boolean functions~\cite{BBDS}.

We lift the remaining operators (except Kleene star) to their list
equivalents: union becomes list append (written $\texttt{+\!+}$);
fusion product is lifted to lists instead of sets, negation is
implemented using list difference ($\setminus$), and the modal
operator lifts the alternative relation over lists of
strings\footnote{Figure~\ref{fig:list-interp} depicts this using the
  list comprehension notation, which is analogous to set builder
  notation, except that it is written using square brackets. Element
  order is evoked by the keyword \texttt{for}, rather than using the
  unordered $\forall$.}. The only caveat to these direct
interpretations is that we restrict the operators to have size $\leq
n$, denoted as $l|_n$ for a list of guarded strings $l$.

@@

%% The lists we use here are \emph{lazy} (as opposed to \emph{strict}), which
%% broadly means that computation is delayed until the value is needed. This allows
%% us to avoid computing large, unnecessary iterations.

@@

Sentences are parsed with a chart parser for categorial grammar.  The
semantics for complex phrases are obtained by application of semantic
translations, accompanied by beta reductions that eliminate all
lambdas in logical forms for clauses.  In consequence, the semantic
term translating a sentence is an Epik term.  Such a term designates a
set of possible words (guarded strings).  By way of example,
\exi{three}{a} is an English sentence with conjunction and several
levels of clausal embedding.  Using the grammar and parser, the
sentence is mapped to the term in \exi{three}{b}. Then \exi{three}{c}
shows a simplified logical from constructed form \exi{three}{b} using
simply logical equivalences. This term is compiled in an
implementation of the finite state calculus to a finite state machine
with six states and eight arcs, which accepts a countably infinite set
of worlds. \ehc{Is this right anymore? the formula has changed
  drastically}. In this way the methodology ``directly'' represents
the set of worlds denoted by \exi{three}{a}.

@@

\begin{examples} \label{three}
\itema It's tails and Amy knows that Bob knows that Amy knows whether it's heads.
% (_*;t) & ~ <amy> ~ ~ <bob> ~((~ (<amy> ~ (_*;h))) + ~ (<amy> (_*;h)) )
\itemb \( \displaystyle \bullet^*t \wedge \AKBKAKWH \)
\itemc \(\bullet^*t \wedge \K_{a}(\K_{b}(\K_{a} \bullet^* h + \K_{a} \neg \bullet ^* h))\)
\itemd Amy knows that it's tails.
\iteme \(
\bullet ^*t \wedge \neg \Diamond_{a} \neg \bullet^*t~~~~~~(\equiv \K_a \bullet^* t)
\)

@@

\section{Discussion}
The methodology presented here is designed for use in research in
linguistic semantics, and for education at the level of a second
graduate course in formal semantics, covering intensionality.  There
are straightforward extensions to additional semantic phenomena, such
as tense and perfective aspect as in \exi{extension}{a}, and the combination of
metaphysical modality and prospective aspect in \exi{extension}{b}.
